A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-05-08 graph_syntax.tex
Major update

\section{Requirements for New Graph Syntax}
  \label{s:requirements}
  The goal of the new graph--based syntax presented here is to enable the general structure of an MDAO problem to be described independently of any solution information, 
  while still being able to accomodate the more specific case when a solution 
  strategy is applied. In order to achieve this goal, 
  the graph syntax needs to accomodate a number of MDAO problem constructs: 
  \begin{itemize}
    \item Analysis tools and their interconnections
    \item Design variables, objectives, and constraints
    \item Local and global information
    \item Coupling between analyses
    \item Multi-fidelity analyses
  \end{itemize}

  The new syntax is intended to represent three phases of the design problem 
  formulation process. In the initial problem definition phase, the specific 
  analysis tools and design goals are identified. Next, a single formal problem 
  formulation is identified that specifies design variables, constraints, 
  objectives, analysis tools, and all other elements required to represent 
  the overall MDAO problem. Lastly, a specific procedure for solving the problem 
  is selected, e.g. selecting an MDAO optimization architecture. Using 
  the proposed graph syntax, these phases can be represented with the following 
  three graphs:

  \begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem Graph (FPG)
    \item Problem Solution Graph (PSG)
  \end{itemize}

  The \emph{maximal connectivity graph} represents the first phase of the problem formulation with all 
  analysis codes being considered and all possible connections between them also present. The second graph 
  is the \emph{fundamental problem graph}, which is the smallest possible graph 
  that still fully defines a given problem formulation that could in principle be solved by an MDAO technique. Finally, a \emph{specific problem formulation} 
  may be represented by including additional edges and nodes to represent the 
  solution strategy being employed to solve the problem.

  The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
  The tree diagram demonstrates the fact that it is generally possible to obtain 
  multiple FPGs from a single maximal connectivity graph. This  may correspond to 
  different down--selections of analysis codes, different connections between them, 
  or both. Each down--selection shrinks the number of possible FPGs that could be reached 
  until only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
  different solution strategies. If you consider the size of a graph to be the sum of all of its
  edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
  the FPG is obtained from the MCG by removing nodes and edges, 
  and the PSG is obtained from the FPG by adding nodes and edges.

  \begin{figure}[htb!]
    \centering
    \subfigure[number of possible graphs]{
    \includegraphics[width=2.0in]{images/tree}
    \label{f:tree}
    }
    \subfigure[graph size]{
    \includegraphics[width=2.0in]{images/hourglass}
    \label{f:hourglass}
    }
  \caption{The relationship between the MCG, FPG, and PSG.}
  \end{figure}

\section{Syntax Definition}
  \label{s:syntax definition}
  In this section, we present the necessary graph theoretic fundamentals to 
  construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
  the edges directed out from $v$ are denoted by $E(v)$ and the edges directed into $v$ are given 
  by $E^{-1}(v)$. 
  If $E$ is not a one--to--one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

  A \emph{path} $P=(V,E)$ from $x_0$ to $x_k$ in graph $G$ is a subgraph of $G$ 
  with $V = \{x_0,x_1,\ldots,x_k\}$ and $E = \{(x_0,x_1),(x_1,x_2),\ldots,(x_{k-1},x_k)\}$.
  A \emph{reverse path} $P_R$ in graph $G$ is a path on $R$, where $R$ is the 
  reverse graph of $G$ obtained by switching the orientation of every edge.

@@

\subsection{Node and Edge Types}
  \label{ss:types}
  We now define different types of nodes and edges with specific properties to compose graphs used to represent an MDAO problem. 
  There are three node types:  
  \begin{description}
    \item[variable:] represents scalar or array data
    \begin{equation} 
      V_{\txt{var}} = \{v \st t_{\txt{node}}(v) = \txt{`variable'}\}
    \end{equation}

    \item[model:] responsible for mapping inputs to outputs
    \begin{equation} 
      V_{\txt{model}} = \{v \st t_{\txt{node}}(v) = \txt{`model'}\}
    \end{equation}

    \item[driver:] control logic capable of managing iteration
    \begin{equation} 
      V_{\txt{driver}} = \{v \st t_{\txt{node}}(v) = \txt{`driver'}\}
    \end{equation}

  \end{description}

  In addition to the three node types, there are two edge types:

  \begin{description}
  \item[connection edge:] Exchange of information between two nodes. These edges 
  can either be fixed (cannot be removed from graph) or free (can be removed). 
    \begin{equation} 
      E_{\txt{conn}} = \{e \st t_{\txt{edge}}(e) = \txt{`connection'}\}
    \end{equation}
  \item [driven edge:] Passing of information from a driver node to a 
    variable node. A single variable node can have many incomming driven edges. These edges are 
    free and can be added or removed from the graph.
     \begin{equation} 
       E_{\txt{driven}} = \{e \st t_{\txt{edge}}(e) = \txt{`driven'}\}
     \end{equation} 
  \end{description}

  [[Insert legend for node and edge types]]

  In Alexandrov and Lewis's REMS syntax only two node types (variable and function) 
  and one edge type are included \cite{alexandrov2004}. We have chosen to adopt 
  the terminology ``model'' node instead of ``function'' node to be consistent 
  with modern MDAO framework terminonlogy. The present work adds one additional 
  node types and one additional edge type to the syntax to allow descriptive
  graphs for all three phases of the design problem formulation process. In addition, we introduce the 
  concept of fixed vs. free connection edges to represent the
  static relationship between variable and model nodes from a single analysis, 
  vs the flexible relationship between variables from different analyses.

\subsection{Rules for Nodes and Edges}
  \label{ss:rules}
  A rule set is provided for the usage of these nodes and edges.
  The driver node and the driven edge are allowed to be present only in PSGs.
  Other node and edge types can be present in any of the three graph types, 
  subject to the following restrictions: 
  \begin{enumerate}
  \item A model node may have only one edge directed to or from another model node.
  \item A model node may have only fixed connection edges directed in or out.
  \item A model node must have at least one edge directed in and at least one edge 
    directed out.
  \item If a variable node has an outgoing edge to a model node, then it may not have other outgoing edges.
  \end{enumerate}

\subsection{indegree and outdegree}
  \label{s:indegree-outdegree}
  The \emph{indegree} of a node is the number of connection edges directed in and 
  is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
  is the number of connection edges directed out and it is denoted as $\txt{deg}^+(v)$.
  The degree of a given node is a function only of the connection edges 
  attached to it. The number of driven edges is not relevant because any number 
  of drivers could be involved in different parts of an solution process. For 
  example in a sequential optimization in which a global optimizer is used first 
  followed by a gadient based optimizer second to refine global optimizer 
  result, design variables would have incomming driven edges from both optimizers. 
  These driven edges would not affect the indegree of those variable nodes.

  Specifially for variable nodes we also define the \emph{upper indegree limit}

  \begin{equation}
  \txt{deg}_u^-(v):V \to \mathbb{N}
  \end{equation} 
  and the \emph{lower indegree limit} as
  \begin{equation}
  \txt{deg}_l^-(v):V \to \mathbb{N}.
  \end{equation}
  These user-specified limits govern the number of connection edges that may 
  be directed into a variable node for a valid graph. Consider a variable 
  node $v$ with $\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, $v$ 
  must have exactly one incomming explicit edge or the graph is deemed invalid. 
  Two possiblities exist for voilating these limit conditions:

  \begin{description}
    \item[hole: ] The number of incomming edges is less than the lower indegree limit:
      \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation}
      A hole represents a lack of information being supplied to the node.
    \item[collision: ] The number of incomming edges is greater than $ \txt{deg}_u^-(v)$. 
      \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \label{e:collision}\end{equation}
  \end{description}

  The presence of holes and collisions in a graph represent issues that will give
  rise to an invalid problem formulation. An algorithm for detecting and 
  managing these violations if proposed in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG}.

\section{Graph Representation of MDAO Constructs}
\label{s:graph representation}
In this section we present our graph syntax applied to the description of MDAO 
Problems. We make use of the Sellar problem, described in 
Eq. \ref{eqn:simple_fpf} as an example to illustrate the implementation of the syntax. 
The graph representation of the Sellar problem, using the new stynax, is 
presented in Fig. \ref{f:sellar_graph_full}. In this section sub-graphs or 
slight modifications of the full graph will be used to demonstrate the 
important MDAO constructs described by this graph syntax.

\begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=4in]{images/sellar_graph_full}
    \end{center}
    \vspace{-10pt}
\caption{Graph of the Sellar problem formulation}
\label{f:sellar_graph_full}
\end{figure}

\subsection{Analysis Blocks and Connections}
\label{ss:analysis blocks and connections}
Analysis tools take in a number input variables and then calculate 
the values for their respective outputs. We represent this process
by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Each analysis block is 
denoted by:

\begin{equation}
  A_i=(V_{A_i},E_{A_i}),\ for\ i\in I, 
\end{equation}

Where $m$ is the number of analysis tools represented in the graph 
and $I=\{1,2,\ldots,m\}$.

Within a given analysis block, variable node represents a single input or 
output and is connected to a single model node via a fixed connection edge. Note that 
in Fig.~\ref{f:analysis block}, the analysis block includes two model nodes, 
with a single edge connecting them. This edge represents the necessary 
calculations to map given inputs to the outputs of the anlysis. This calculation edge provides the opportunity 
to encode computational cost or other characteristics of the analysis code as 
a weight in a weighted graph formulation. If associating a weight is 
unnecessary, the calculation edge and second model node may be omitted and all input and output 
variables can be connected to a single model node for a given analysis block. 
Since all the edges within analysis blocks are fixed, the blocks themselves are 
fixed sub-graphs within the overall MDAO problem graph. The connectivity of 
nodes and edges in an analysis block cannot be altered for the purposes of MDAO 
problem formulation; however, analysis block sub-graphs can be added or removed 
from the MDAO problem graph as needed.

\begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=4in]{images/analysis_block}
    \end{center}
    \vspace{-10pt}
\caption{Example analysis block. The each node type and edge type is labeled in italics and annotated parenthetically.}
\label{f:analysis block}
\end{figure}

Variable nodes in an analysis block can be distinguished as either an input or 
an output by the manner in which they are connected. As shown in Fig.~\ref{f:analysis block}, 
inputs are represented as variable nodes that have an outgoing edge into a model 
node. Conversely, outputs represented as any variable nodes that have an incoming 
edge from a model node.

MDAO problems require that information be passed between sets of analyses. When 
information from one output of an analysis block is passed, or connected, to an 
input of another analysis block, a new connection edge is added between the two 
variable nodes involved in the exchange. These connection edges are free, and unlike the edges 
within an analysis block, the edges can be added or removed depending on the needs
of a given design problem. In other words, free connection edges are created by 
engineers linking an output of one tool to an input of another one. It is 
allowable for a single output to have outgoing connection edges to multiple 
downstream inputs.

\subsection{Design Variables}
Design variables are the input variables that are free to be 
changed by the designer or by an optimizer in an MDAO study. In a graph design
variables will be a set of variable nodes, $V_{dv}$, identified as follows:

\begin{equation}
  V_{\txt{dv}} = \{V_{\txt{var}} \st \txt{deg}_l^-(v)=0\}
\end{equation}

For the Sellar problem, there are three design variables: $x_1^*$, $z_1^*$, and 
$z_2^*$. In Fig. \ref{f:designvars} all variable from the Sellar problem are 
annotated with their lower indegree limit and the three design variables are identified 
in bold.

\begin{figure}[htb!]
  \begin{center}
    [Graph Here....]
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Sellar problem graph with lower indegree limits for all variables \label{f:designvars}}
\end{figure}

In Fig. \ref{f:designvars}, none of the design variable nodes have any incomming 
edges. In general, in a MCG or a FPG, design variables will 
show up as nodes without any incomming edges. For a valid PSG, design variables 
could have one or more incomming driven edges representing a value determined by 
a optimizer.

Note the although there are no incomming edges, design variables are not holes 
in the graph as defined by Eq. \ref{e:hole}. Since they have $\txt{deg}_l^-(v)=0$ 
having no incomming connection edges does not violate their lower indegree limit. 
However, the concept of holes can be useful for identifying potential design variables
given an arbitrary MCG. If you first set $\txt{deg}_l^-(v)=1$ for all variable nodes, 
and then find the resulint holes, you will have a list of variable nodes that 
could potentially be design variables. Some variables will be true design variables, 
while others might represent missing information in the problem formulation. For 
instance, if there is a hole for a drag input of an aircraft mission analysis code 
then this is not likely a good choice for a design variable. So it is up to the designer to 
examine each hole and decide if it's an appropriate design variable. If 
the variable should become a design variable then set $\txt{deg}_l^-(v)=0$. 
If a variable should not be a design variable, then leave it as a hole which 
needs to be addressed in a manner discussed in section \ref{ss:obtaining FPG}.

\subsection{Objectives and Constraints}
\label{ss:objectives and constraints}
Objectives and constraints are constructs that need to be identified when 
describing and MDAO problem. In the case of objective functions a single output 
value generated by an analysis block could be selected, but commonly multiple 
values are assembeled together to form a composite objective function. 
Generally both objective and constraint functions accept a set of inputs and map 
them to an output value of significance to the overall design problem.

The operations of implementing composite objective and constraint functions, 
although typically simple, are effectively the identical to the task performed by an 
analysis block. A composit constraint or objective function can therfore 
be represented as an additional analysis block within the graph, with its own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience, it is useful to distinguish between analysis 
blocks corresponding to analysis codes and those that arise from the addition of 
objectives or constraints to the graph. Therefore, we denote an \emph{expression block}, as:

\begin{equation}
  F_j=(V_{F_j},E_{F_j}),\ for\ j\in J, 
\end{equation}

Where $m$ is the number of expressions in the graph
and $J=\{1,2,\ldots,m\}$. For the sake of consistency of naming conventions
$F_j$, $G_j$, and $H_j$ are used to deonte expression blocks for objectives, inequality
constraints, and equality constraints. If the relevant 
objective and constraint values are direct results from other analysis blocks 
and composite functions are not needed, then expression blocks may be omitted 
from the MDAO problem graph. Fig. \ref{f:obj-cons} highlights the 
objective and constraint blocks from the Sellar problem.

\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/obj_const_graph}
  \end{center}
  \caption{Notional graph with and objective and constraint nodes \label{f:obj-cons}}
\end{figure}

\subsection{Local vs Global Nodes}

  The set of local nodes, $V_\txt{l}$, is the collection of 
  nodes with exactly one outgoing connection edge.

  \begin{equation}
    V_{\txt{l}} = \{V_{\txt{var}} \st \txt{deg}^+(v)=1\}
    \label{e:local_definition}
  \end{equation}

  In section \ref{s:specific vs fundamental} variable nodes that are part of 
  an analysis block are inherently local and similarly model nodes 
  are also local. Following the definition in Eq. \ref{e:local_definition}
  we can now see why. Variable nodes connected to model nodes are only allowed to 
  have one outgoing edge, and hence will always by definition be local variables.

  The set of global nodes, $V_\txt{g}$, is the collection of all nodes with more than 
  one outgoing connection edge.

  \begin{equation}
    V_{\txt{g}} = \{V_{\txt{var}} \st \txt{deg}^+(v)>1\}
    \label{e:global_definition}
  \end{equation}

  Follwoing our definition any variable node that had multiple 
  outgoing edges, is global. In Fig. \ref{f:sellar_graph_full} 
  the variable nodes for $x_1^*$, $z_1^*$, and $z_2^*$ have multiple outgoing 
  edges and would be considered global.

  The variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem, $x_1$ is considered a local 
  variable because it directly affects only analysis block $A_1$. By expanding
  $F$ into an expression block with its own inputs, then $F$ is assigned its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global $x_1^*$ 
  node in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the different blocks. Interestingly, the classification of $x_1^*$ 
  as a global variable fits nicely within the structure of the Colaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for setting up 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In that case a global 
  variable is created, and the analysis block again is again assigned an 
  independent local copy of it. Given our definition of global vs local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a given variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  result in the correct structure for CO automatically.

  Analysis and expression blocks can themselves be local or global, but the 
  criteria for determining this is slightly different than for variable nodes. 
  In this case, you consider the set of connection edges for the block
  (e.g. from all the variable nodes within the block) as a set. Additionally, 
  you must consider both incoming and outgoing edges, where for variables 
  only outgoing connection edges were relevant.

>>>>

\section{Graph-Based Syntax Definition and Formalisms}
  \label{s:syntax definition}
  In this section we present our graph-based syntax, beginning with the necessary graph theoretic fundamentals to 
  construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this paper is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2-element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. An edge $e$ = $(x,y)$ may be referred to simply as $xy$. The edges directed out from node $v$ are denoted by $E(v)$ and the edges directed into $v$ are given 
  by $E^{-1}(v)$. 
  If $E$ is not a one-to-one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

  A \emph{path} $P=(V,E)$ from $x_0$ to $x_k$ in graph $G$ is a subgraph of $G$ 
  with $V = \{x_0,x_1,\ldots,x_k\}$ and $E = \{(x_0,x_1),(x_1,x_2),\ldots,(x_{k-1},x_k)\}$. 
Path $P$ is a \emph{cycle} if $x_0 = x_k$.
  A \emph{reverse path} $P_R$ in graph $G$ is a path on $R$, where $R$ is the 
  reverse graph of $G$ obtained by switching the orientation of every edge.

@@

\subsection{Node and Edge Types}
  \label{ss:types}
Along with these fundamentals of graph theory, we categorize nodes and edges into distinct types in order to provide an intuitive approach to MDAO problem formulation. The three node types are
  \begin{description}
    \item[\bf{variable:}] represents scalar or array data, inputs and outputs,
    \item[\bf{model:}] represents the computation peformed by analysis tools,
    \item[\bf{driver:}] control logic capable of managing iterations (PSG),
  \end{description}
and the three edge types are
%$V_\txt{var}$, $V_\txt{model}$, $V_\txt{driver}$
  \begin{description}
  \item[\bf{connection edge:}] exchange of information external to analysis tools,
  \item[\bf{model edge:}] exchange of information internal to analysis tools,
  \item[\bf{driven edge:}] passing of information from a driver node to a variable node (PSG).
  \end{description}
Figure \ref{f:sellar types} demonstrates the usage of these node and edge types via the Sellar problem. Model nodes are indicated by squares, variable nodes are indicated by circles, connection edges are indicated by dotted lines, and model edges are indicated by solid lines.
\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=4.0in]{images/sellar_types}
  \end{center}
  \caption{Sellar problem represented as a graph.}
\label{f:sellar types}
\end{figure}

% ((JG: A single variable node can have many incomming driven edges. These edges are 
%    free and can be added or removed from the graph.))
%
%  [[Insert legend for node and edge types]] ((DP: I'm not sure what the legend is supposed to be))

  A rule set is provided for the usage of these node and edge types to provide a structure to the graph-based representation.
  The driver node and the driven edge are allowed to be present only in PSGs, whereas the other node and edge types can be present in any of the three graph types, 
  subject to the following restrictions: 
  \begin{enumerate}
  \item A model node may have only one edge directed to or from another model node.
  \item A model node may have only model edges directed in or out.
  \item A model node must have at least one edge directed in and at least one edge 
    directed out.
  \item If a variable node has an outgoing model edge, then it may not have other outgoing edges.
  \item If a variable node has an incoming model edge, then it may not have any additional incoming edges.
  \end{enumerate}

  In Alexandrov and Lewis's REMS syntax only two node types (variable and function) 
  and one edge type are included \cite{alexandrov2004}. We have chosen to adopt 
  the terminology ``model'' node instead of ``function'' node to be consistent 
  with modern MDAO framework terminology. The present work adds one additional 
  node type and two additional edge types to the syntax to allow descriptive
  graphs for all three phases of the design problem formulation process indicated in Figs.~\ref{f:tree} and \ref{f:hourglass}.
\begin{comment}
In addition, we introduce the 
  concept of fixed vs. free connection edges to represent the
  static relationship between variable and model nodes from a single analysis, 
  as opposed to the flexible relationship between variables from different analyses. 
\end{comment}

\subsection{Analysis Blocks}
\label{ss:analysis blocks}
Analysis tools take in a set of input variables and then calculate 
the values for their respective outputs. We represent this process
by a digraph called an \emph{analysis block}; a notional analysis block is shown in Fig.~\ref{f:analysis block}. 
%Each analysis block is a directed graph denoted as $ A=(V_{A},E_{A})$
As indicated in this figure, analysis blocks comprise three sets of nodes representing the distinct local inputs, the analysis (computation), and the distinct local outputs. The local input and local outputs nodes are variable nodes, while the nodes representing the analysis are model nodes. All of the edges within the analysis block are model edges, and are considered fixed to the analysis block. This graph structure demonstrated by Fig.~\ref{f:analysis block} satisfies the rules listed in Sec.~\ref{ss:types}. Reciprocally, the rules ensure that analysis tools are represented via this analysis block structure.
\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=3.0in]{images/analysis_block}
    \end{center}
    \vspace{-10pt}
\caption{Example analysis block with node types labeled.}
\label{f:analysis block}
\end{figure}

The model edge connecting the two model nodes represents the necessary calculations to map the inputs to the outputs of the analysis. 
This edge also provides the opportunity to encode computational cost or other characteristics of the analysis code as a weight in a weighted graph formulation. 
Since all the edges within analysis blocks are fixed, the blocks themselves are fixed sub-graphs within the overall MDAO problem graph. 
The connectivity of nodes and edges in an analysis block cannot be altered for the purposes of MDAO problem formulation; 
however, analysis block sub-graphs can be added or removed from the MDAO problem graph as needed.
Variable nodes in an analysis block can be distinguished as either an input or an output by the manner in which they are connected. 
As shown in Fig.~\ref{f:analysis block}, inputs are represented as variable nodes that have an outgoing edge into a model node, and outputs are represented as variable nodes that have an incoming edge from a model node.

%MDAO problems require that information be passed between sets of analyses ((DP: needs better transition)).
%For this purpose, connection edges are placed to link analysis blocks
% 
%When information from one output of an analysis block is passed to an 
%input of another analysis block, a new connection edge is added between the two 
%variable nodes involved in the exchange. 
%These connection edges are free, and unlike the edges within an analysis block, the edges can be added or removed depending on the needs of a given design problem. 
%In other words, free connection edges are created by engineers linking an output of one tool to an input of another one. 
%It is allowable for a single output to have outgoing connection edges to multiple downstream inputs.

\subsection{Objectives, Constraints, and Global Inputs}
\label{ss:objectives and constraints}
Along with analysis tools, objectives, constraints, and global inputs also need to be represented.
In the case of objective functions, a single output value generated by an analysis block could be selected, but commonly multiple values are assembeled together to form a composite objective function. 
Generally, both objective and constraint functions accept a set of inputs and map them to an output value of significance to the overall design problem.

The operations of implementing composite objective and constraint functions, although typically simple, are effectively identical to the task performed by an analysis block. 
A composite constraint or objective function can therefore be represented as an analysis block within the graph with its own input and output variable nodes. 
Although fundamentally no different than an analysis block, for clarity and convience, it is useful to distinguish between analysis blocks corresponding to analysis codes and those that arise from the addition of objectives or constraints to the graph. 
Therefore, we use an \emph{expression block} to represent an objective or constraint. 
These expression block graphs follow the same structure as analysis blocks presented in Sec.~\ref{ss:analysis blocks}.

Inputs to multiple analysis tools are considered global inputs. These are represented in the graph syntax by single variable nodes refered to as global inputs---the definition of global with respect to a node will be defined subsequently.
%For the sake of consistency of naming conventions $F_j$, $G_j$, and $H_j$ are used to deonte expression blocks for objectives, inequality constraints, and equality constraints. 
%If the relevant objective and constraint values are direct results from other analysis blocks and composite functions are not needed, then expression blocks may be omitted from the MDAO problem graph. 
Figure \ref{f:obj-cons} demonstrates the use of global inputs (black circles), analysis blocks (dotted boxes), and expression blocks (dashed boxes), for the Sellar problem graph.
\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=4.0in]{images/sellar_obj_and_cons}
  \end{center}
  \caption{Sellar problem graph indicating global inputs (black circles), analysis blocks (dotted boxes), and expression blocks (dashed boxes).}
\label{f:obj-cons}
\end{figure}

\subsection{Indegree and Outdegree Limits}
  \label{s:indegree-outdegree}
	To address the MDAO notion of design variables, we first introduce the concept of the degree of a node.
  In a digraph, \emph{indegree} of a node is the number of edges directed in and 
  is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
  is the number of edges directed out and it is denoted as $\txt{deg}^+(v)$ \cite{Diestel2010}.
  In this paper, the indegree and outdegree of a given node refer only to the number of the connection edges 
  directed into or out of the node, respectively. 
%The number of driven edges is not considered relevant because any number 
%  of drivers could be involved in different parts of a solution process. For 
%  example, in a sequential optimization in which a global optimizer is used first 
%  followed by a gradient-based optimizer second to refine global optimizer 
%  result, design variables would have incomming driven edges from both optimizers. 
%  These driven edges would not affect the indegree of those variable nodes.

  We may now define the \emph{upper indegree limit} 
  \begin{equation}
  \txt{deg}_u^-(v):V \to \mathbb{N}
  \end{equation} 
  and the \emph{lower indegree limit} as
  \begin{equation}
  \txt{deg}_l^-(v):V \to \mathbb{N}.
  \end{equation}
  These user-specified limits govern the number of connection edges that may 
  be directed into a node for a valid graph; of course, $\txt{deg}_l^-(v) \leq \txt{deg}_u^-(v)$ must be satisfied. 
For example, consider a variable 
  node $v$ with $\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, $v$ 
  must have exactly one incoming connection edge or the graph is deemed an invalid problem formulation.

\section{MDAO Constructs Derived from the Graph-Based Syntax}
\label{s:graph representation}
In Sec.~\ref{s:syntax definition} we presented the structure of the graph-based representation. We now discuss how these formalisms provide the remaining MDAO problem constructs identified in Sec.~\ref{s:requirements}.

%To begin, we make use of the Sellar problem, described in 
%Eq. \ref{eqn:simple_fpf} as an example to illustrate the implementation of the syntax. 
%The graph representation of the Sellar problem, using the new stynax, is 
%presented in Fig. \ref{f:sellar_graph_full}. In this section sub-graphs or 
%slight modifications of the full graph will be used to demonstrate the 
%important MDAO constructs described by this graph syntax.
%\begin{figure}[htb!]
%    \begin{center}
%    \includegraphics[width=4in]{images/sellar_graph_full}
%    \end{center}
%    \vspace{-10pt}
%\caption{Graph of the Sellar problem formulation}
%\label{f:sellar_graph_full}
%\end{figure}

\subsection{Design Variables, Holes, and Collisions}
Design variables are input variables whose values are free to be 
changed by the designer or by an optimizer in an MDAO study. In the graph-based syntax, a node corresponds to a design variable when the indegree limit of the node is prescribed as zero.

If the indegree limits are violated by the number of incoming connection edges (the indegree of the node), the graph is regarded as an invalid problem formulation. There are two classifications for these violations:
  \begin{description}
    \item[\bf{hole:}] The number of incoming edges is less than the lower indegree limit:
      \begin{equation} 
\txt{deg}^-(v) < \txt{deg}_l^-(v). 
\label{e:hole} 
\end{equation}
      A hole represents a lack of information being supplied to the variable node. This implies that the analysis tool being represented by the analysis block would not be capable of producing outputs.
    \item[\bf{collision:}] The number of incoming edges is greater than the upper indegree limit. 
      \begin{equation} 
\txt{deg}^-(v) > \txt{deg}_u^-(v). 
\label{e:collision}
\end{equation}A collision represents redundant information being supplied to the variable node from two or more sources. This conflict implies an ambiguity as to which information is to be used as an input for the analysis tool.
  \end{description}

%For the Sellar problem, there are three design variables: $x_1^*$, $z_1^*$, and 
%$z_2^*$. In Fig. \ref{f:designvars} each variable is 
%annotated with the lower indegree limit, and the three design variables are identified 
%in bold((DP:check)). 
%
%\begin{figure}[htb!]
%  \begin{center}
%    [Graph Here....]
%    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
%  \end{center}
%  \caption{Sellar problem graph with lower indegree limits for all variables \label{f:designvars}}
%\end{figure}
%
%In Fig. \ref{f:designvars}, the design variable are void of incoming edges. 
%For a valid PSG, design variables could have one or more incoming driven edges representing a value determined by an optimizer.

  The presence of holes and collisions in a graph represent conflicts that will give
  rise to an invalid problem formulation. It is expected that a maximal connectivity graph will posess these conflicts, whereas a fundamental problem graph will not.

%However, if the graph remains unchanged while the user changes the lower indegree limit of the node to now be zero, then the node becomes a design variable.

%In some cases, the distinction between a hole and a design variable is rather subtle. Both have no incoming connection edges, but 
%
%
%Note that although there are no incoming edges, design variables are not holes 
%in the graph as defined by Eq. \ref{e:hole}. Since they have $\txt{deg}_l^-(v)=0$ 
%having no incoming connection edges does not violate their lower indegree limit. 
%However, the concept of holes can be useful for identifying potential design variables
%given an arbitrary MCG. If you first set $\txt{deg}_l^-(v)=1$ for all variable nodes, 
%and then find the resulint holes, you will have a list of variable nodes that 
%could potentially be design variables. Some variables will be true design variables, 
%while others might represent missing information in the problem formulation. For 
%instance, if there is a hole for a drag input of an aircraft mission analysis code 
%then this is not likely a good choice for a design variable. So it is up to the designer to 
%examine each hole and decide if it's an appropriate design variable. If 
%the variable should become a design variable then set $\txt{deg}_l^-(v)=0$. 
%If a variable should not be a design variable, then leave it as a hole which 
%needs to be addressed in a manner discussed in section \ref{ss:obtaining FPG}.

\subsection{Local vs Global Nodes ((recommend omission))}

  In this graph-based syntax, a variable node is considered \emph{local} if it has exactly one outgoing connection edge or \emph{global} if it has more than one outgoing connection edge. Therefore, a variable node may be used to represent a global input if it has at least two outgoing connection edges. Otherwise, the input is formulated as a design variable that is local to a specific analysis block.
%  The set of local nodes, $V_\txt{l}$, is the collection of 
%  nodes with exactly one outgoing connection edge. 
%
%  \begin{equation}
%    V_{\txt{l}} = \{V_{\txt{var}} \st \txt{deg}^+(v)=1\}
%    \label{e:local_definition}
%  \end{equation}
%  In section \ref{s:specific vs fundamental} variable nodes that are part of 
%  an analysis block are inherently local and similarly model nodes 
%  are also local. Following the definition in Eq. \ref{e:local_definition}
%  we can now see why. Variable nodes connected to model nodes are only allowed to 
%  have one outgoing edge, and hence will always by definition be local variables.

%  The set of global nodes, $V_\txt{g}$, is the collection of all nodes with more than 
%  one outgoing connection edge. (())
%
%  \begin{equation}
%    V_{\txt{g}} = \{V_{\txt{var}} \st \txt{deg}^+(v)>1\}
%    \label{e:global_definition}
%  \end{equation}
